Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-10-24 paper/old.tex

% supplemental tables
\newcommand{\questions}{1}
\newcommand{\topics}{2}
\newcommand{\matchTab}{3}

% supplemental figures
\newcommand{\topicWordWeights}{1}
\newcommand{\topicWeights}{2}
\newcommand{\individualKnowledgeMapsA}{3}
\newcommand{\individualKnowledgeMapsB}{4}
\newcommand{\individualKnowledgeMapsC}{5}
\newcommand{\individualLearningMapsA}{6}
\newcommand{\individualLearningMapsB}{7}
\newcommand{\ldaVsBERT}{8}
\newcommand{\jaccard}{9}   % word match figure

@@

Designing and building procedures and tools for mapping out knowledge touches
on deep questions about what it means to learn. For example, how do we acquire
conceptual knowledge? Memorizing course lectures or textbook chapters by rote
can lead to the superficial \textit{appearance} of understanding the underlying
content, but achieving true conceptual understanding seems to require something
deeper and richer. Does conceptual understanding entail connecting newly
acquired information to the scaffolding of one's existing knowledge or
experience~\citep{BlayEtal06,CaraMaho03, ConsEtal16, DeacEtal04, SimoEtal04,
HuebWill18}? Or weaving a lecture's atomic elements (e.g., its component words)
into a structured network that describes how those individual elements are
related~\citep{LeeChen22, vanPEtal21}? Conceptual understanding could also
involve building a mental model that transcends the meanings of those
individual atomic elements by reflecting the deeper meaning underlying the
gestalt whole~\citep{Kint70, Macl05, ScotEtal07, TulcEtal23}.

The difference between ``understanding'' and ``memorizing,'' as framed by
researchers in education, cognitive psychology, and cognitive neuroscience
(e.g.,~\citealp{Kato40, Gall00, ScotEtal07, HallGree08, Macl05}), has profound
analogs in the fields of natural language processing and natural language
understanding. For example, considering the raw contents of a document (e.g.,
its constituent symbols, letters, and words) might provide some clues as to
what the document is about, just as memorizing a passage might provide some
ability to answer simple questions about it. However, text embedding models
(e.g.,~\citealp{LandDuma97, DeerEtal90, BleiEtal03, BleiLaff06, MikoEtal13a,
CerEtal18, BrowEtal20, ViswEtal17}) also attempt to capture the deeper meaning
\textit{underlying} those atomic elements. These models consider not only the
co-occurrences of those elements within and across documents, but (in many
cases) also patterns in how those elements appear across different scales
(e.g., sentences, paragraphs, chapters, etc.), the temporal and grammatical
properties of the elements, and other high-level characteristics of how they
are used~\citep{Mann20, Mann21a}. According to these models, the deep
conceptual meaning of a document may be captured by a feature vector in a
high-dimensional representation space, wherein nearby vectors reflect
conceptually related documents. A model that succeeds at capturing an analogue
of ``understanding'' is able to assign nearby feature vectors to two
conceptually related documents, \textit{even when the specific words contained
in those documents have very little overlap}. In this way, ``concepts'' are
defined implicitly by the model's geometry~\citep[e.g., how the embedding coordinate
of a given word or document relates to the coordinates of other text
embeddings;][]{PianHill22}.

Given these insights, what form might a representation of the sum total of a
person's knowledge take (speculatively)? First, we might require a means of
systematically describing or representing the nearly infinite set of possible
things a person could know. Second, we might want to account for potential
associations between different concepts. For example, the concepts of ``fish''
and ``water'' might be associated in the sense that fish live in water. Third,
knowledge may have a critical dependency structure, such that knowing about a
particular concept might require first knowing about a set of other concepts.
For example, understanding the concept of a fish swimming in water first
requires understanding what fish and water \textit{are}. Fourth, as we learn,
our ``current state of knowledge'' should change accordingly. Learning new
concepts should both update our characterizations of ``what is known'' and also
unlock any now-satisfied dependencies of those newly learned concepts so that
they are ``tagged'' as available for future learning.

@@

It is important to clarify that although we use topic model-derived embeddings
to \textit{characterize} the conceptual content of the lectures and questions,
we do not claim that the topic model itself \textit{understands} the conceptual
content of the lectures or questions. Rather, we view the topic model as a tool
for capturing the \textit{structure} of the conceptual content of the lectures
and questions in a way that enables us to capture, quantify, and track and
predict participants' knowledge.

@@

    \caption{\textbf{Estimating knowledge about the content presented at each
    moment of each lecture.} \textbf{A. Knowledge about the time-varying
    content of \textit{Four Fundamental Forces}}. Each trace displays the
    weighted proportion of correctly answered questions about the content
    reflected in each moment of the lecture (see \nameref{subsec:traces}),
    using responses from a single quiz (color). The traces are averaged across
    participants. \textbf{B. Average estimated knowledge about \textbf{Four
    Fundamental Forces}.} Each bar displays the across-timepoint average
    knowledge, estimated using the responses to one quiz's questions.
    \textbf{C. Knowledge about the time-varying content of \textit{Birth of
    Stars}}. The panel is in the same format as Panel A, but here the knowledge
    estimates are for the moment-by-moment content of the \textit{Birth of
    Stars} lecture. \textbf{D. Average estimated knowledge about \textit{Birth
    of Stars}.} The panel is in the same format as Panel B, but here the
    knowledge estimates are for the content of the \textit{Birth of Stars}
    lecture. All panels: error ribbons and error bars denote 95\% confidence
    intervals, estimated across participants.}

@@

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should
carry some predictive information about whether the participant is likely to
answer that question correctly or incorrectly. We developed a statistical
approach to test this claim. For each question, in turn, we used
Equation~\ref{eqn:prop} to estimate each participant's knowledge at the given
question's embedding space coordinate, using all \textit{other} questions that
participant answered on the same quiz. For each quiz, we grouped these
estimates into two distributions: one for the estimated knowledge at the
coordinates of \textit{correctly} answered questions, and another for the
estimated knowledge at the coordinates of \textit{incorrectly} answered
questions (Fig.~\ref{fig:predictions}). We then used Mann-Whitney $U$-tests to
compare the means of these distributions of estimated knowledge.

\begin{figure}[tp]
    \centering
    \includegraphics[width=0.7\textwidth]{figs/predict-knowledge-questions}

    \caption{\textbf{Estimating knowledge at the embedding coordinates of
    held-out questions.} Separately for each quiz (panel), we plot the
    distributions of predicted knowledge at the embedding coordinates of each
    held-out correctly (blue) or incorrectly (red) answered question. The
    Mann-Whitney $U$-tests reported in each panel are between the distributions
    of estimated knowledge at the coordinates of correctly versus incorrectly
    answered held-out questions. In the top row (``All questions''), we used
    all quiz questions (from each quiz, for each participant) except one to
    estimate knowledge at the held-out question's embedding coordinate. In the
    middle rows (``Across-lecture''), we used all questions about one lecture
    to estimate knowledge at the embedding coordinate of a held-out question
    about the \textit{other} lecture. In the bottom row (``Within-lecture''),
    we used all but one question about one lecture to estimate knowledge at the
    embedding coordinate of a held-out question about the \textit{same}
    lecture. We repeated each of these analyses using all possible held-out
    questions for each quiz and participant.}

@@

For the initial quizzes participants took (prior to watching either lecture),
participants' estimated knowledge tended to be low overall, and relatively
unstructured (Fig.~\ref{fig:predictions}, left column). When we held out
individual questions and estimated their knowledge at the held-out questions'
embedding coordinates, we found no reliable differences in the estimates when
the held-out question had been correctly versus incorrectly answered. This
``null'' effect persisted when we used \textit{all} of the Quiz 1 questions
from a given participant to predict a held-out question (``All questions''; $U
= 50587, p = 0.723$), when we used questions from one lecture to predict
knowledge at the embedding coordinate of a held-out question about the
\textit{other} lecture (``Across-lecture''; predicting knowledge for held-out
\textit{Four Fundamental Forces Questions} using \textit{Birth of Stars}
questions: $U = 8244, p = 0.184$; predicting knowledge for held-out
\textit{Birth of Stars} questions: $U = 8202.5, p = 0.161$), and when we used
questions from one lecture to predict knowledge at the embedding coordinate of
a held-out question about the \textit{same} lecture (``Within-lecture'';
\textit{Four Fundamental Forces}: $U = 7681.5, p = 0.746$; \textit{Birth of
Stars}: $U = 8125, p = 0.204$). We believe that this reflects a floor effect:
when knowledge is low everywhere, there is little signal to differentiate
between what is known versus unknown.

After watching \textit{Four Fundamental Forces}, estimated knowledge for
held-out correctly answered questions (from the second quiz;
Fig.~\ref{fig:predictions}, middle column) exhibited a positive shift relative
to held-out incorrectly answered questions. This held when we included all
questions in the analysis ($U = 58332, p < 0.001$), when we predicted knowledge
across-lectures (\textit{Four Fundamental Forces}: $U = 6749.5, p = 0.014$;
\textit{Birth of Stars}: $U = 8480, p = 0.016$), and when we predicted
knowledge at the embedding coordinates of held-out \textit{Four Fundamental
Forces} questions using other \textit{Four Fundamental Forces} questions from
the same quiz and participant ($U = 7224, p < 0.001$). This difference did
\textit{not} hold for within-lecture predictions of \textit{Birth of Stars}
knowledge ($U = 7419, p = 0.739$). Again, we suggest that this might reflect a
floor effect whereby knowledge about the content of the \textit{Birth of Stars}
material is relatively low everywhere in that region of text embedding space.

Finally, after watching \textit{Birth of Stars}, estimated knowledge for
held-out correctly answered questions (from the third quiz;
Fig.~\ref{fig:predictions}, right column) was higher for held-out correctly
answered questions than for held-out incorrectly answered questions. This held
when we included all questions in the analysis ($U = 38279, p = 0.022$), when
we carried out across-lecture predictions (\textit{Four Fundamental Forces}: $U
= 6684.5, p = 0.032$; \textit{Birth of Stars}: $U = 6414.5, p = 0.002$), and
and when we carried out within-lecture predictions of held-out \textit{Birth of
Stars} questions using other \textit{Birth of Stars} questions from the same
quiz and participant ($U = 6126, p = 0.006$). However, we found the
\textit{opposite} effect when we carried out within-lecture predictions of
held-out \textit{Four Fundamental Forces} questions using other \textit{Four
Fundamental Forces} questions from the same quiz and participant ($U = 6734, p
= 0.027$). Specifically, held-out correctly answered Quiz 3 questions about
\textit{Four Fundamental Forces} had reliably \textit{lower} estimated
knowledge than held-out incorrectly answered questions. Speculatively, we
suggest that this may reflect participants forgetting some of the \textit{Four
Fundamental Forces} content. If this forgetting happens in a relatively
``random'' way (with respect to spatial distance within the text embedding
space), then it could explain why some held-out questions about \textit{Four
Fundamental Forces} were answered incorrectly, even if questions at nearby
coordinates (i.e., about similar content) were answered correctly. This might
lead our approach to over-estimate knowledge for held-out questions about
``forgotten'' knowledge that particpants answered incorrectly.

That the knowledge estimates derived from the text embedding space reliably
distinguish between held-out correctly versus incorrectly answered questions
(Fig.~\ref{fig:predictions}) suggests that the text embedding space bears at
least some relationship to participants' knowledge. But what does that
relationship look like as we move through the embedding space? For example,
suppose we know that a participant answers a question (at embedding coordinate
$X$) correctly. As we move away from $X$ in the embedding space, how does quiz
performance ``fall off'' with distance? Or, suppose the participant instead
answered that same question \textit{in}correctly. Again, as we move away from
$X$ in the embedding space, how does our confidence that the participant does
\textit{not} know about the content change with distance? We reasoned that,
assuming our space is capturing something about how participants actually
organize their knowledge, conceptual knowledge right around $X$ should be similar
to the participant's knowledge of the content at $X$. And at another extreme, at
some distance (after moving sufficiently far away from $X$), our guesses about
what participants know (based on their response to the question at location $X$)
should be no better than guessing based on their overall proportion of
correctly answered questions—i.e., if $Y$ is very far away from $X$, all we can do
with the participant's response to $X$ is guess that ``their performance on quiz
questions about $Y$ is about equal to their average performance on quiz questions
about any material.''

\begin{figure}[t]
    \centering
    \includegraphics[width=0.8\textwidth]{figs/knowledge-smoothness}

    \caption{\textbf{Quiz performance falls off gradually in text embedding
    space.} \textbf{A. Performance versus distance.} For each participant, for
    each correctly answered question (blue) or incorrectly answered question
    (red), we computed the proportion of correctly answered questions within a
    given distance of that question's embedding coordinate. We repeated this
    analysis for all questions and participants, and separately for each quiz
    (column). The black lines denote the average proportion correct across
    \textit{all} questions included in the analysis at the given distance.
    \textbf{B. Maximum distance for which performance is reliably different
    from the average.} We used a bootstrap procedure
    (see~\nameref{subsec:smoothness}) to estimate the point at which the blue
    and red lines in Panel A reliably diverged from the black line. We repeated
    this analysis separately for correctly and incorrectly answered questions
    from each quiz. \textbf{All panels.} Error ribbons denote
    bootstrap-estimated 95\% confidence intervals.}

    \label{fig:smoothness}
\end{figure}

With these ideas in mind, we asked: conditioned on answering a question
correctly, what proportion of all questions (within some radius, $r$, of that
question's embedding coordinate) were answered correctly? We plotted this
proportion as a function of $r$. Similarly, we could ask, conditioned on
answering a question incorrectly, how the proportion of correct responses
changed with $r$. As shown in Figure~\ref{fig:smoothness}, we found that quiz
performance falls off smoothly with distance, and the ``rate'' of the falloff
does not appear to change across the different quizzes, as measured by the
distance at which performance becomes statistically indistinguishable from a
simple proportion correct score (see~\nameref{subsec:smoothness}). This
suggests that, at least within the region of text embedding space covered by
the questions our participants answered (and as characterized using our topic
model), the rate at which knowledge changes with distance is relatively
constant, even as participants' overall level of knowledge varies across
quizzes or regions of the embedding space.

Knowledge estimates need not be limited to the content of the lectures. As
illustrated in Figure~\ref{fig:knowledge-maps}, our general approach to
estimating knowledge from a small number of quiz questions may be extended to
\textit{any} content, given its text embedding coordinate. To visualize how
knowledge ``spreads'' through text embedding space to content beyond the
lectures participants watched, we first fit a new topic model to the lectures'
sliding windows with (up to)~$k = 100$~topics. Conceptually, increasing the
number of topics used by the model functions to increase the ``resolution'' of
the embedding space, providing a greater ability to estimate knowledge for
content that is highly similar to (but not precisely the same as) that
contained in the two lectures. This change in the number of topics overcame an
undesireable behavior in the UMAP embedding procedure~\citep{McInEtal18},
whereby embedding coordinates for the 15-topic model tended to be ``clumped''
into separated clusters, rather than forming a smooth trajectory through the 2D
space. When we increased the number of topics to 100, the embedding coordinates
in the 2D space formed a smooth trajectory through the space, with
substantially less clumping (Fig.~\ref{fig:knowledge-maps}). We note that we
used these 2D maps solely for visualization; all relevant comparisons, distance
computations, and statistical tests we report above were carried out in the
original 15-dimensional space, using the 15-topic model. Aside from increasing
the number of topics from 15 to 100, all other procedures and model parameters
were carried over from the preceding analyses. As in our other analyses, we
resampled each lecture's topic trajectory to 1~Hz and projected each question
into a shared text embedding space.

@@

We view our work as making several contributions to the study of how people
acquire conceptual knowledge. First, from a methodological standpoint, our
modeling framework provides a systematic means of mapping out and
characterizing knowledge in maps that have infinite (arbitrarily many) numbers
of coordinates, and of ``filling out'' those maps using relatively small
numbers of multiple choice quiz questions. Our experimental finding that we can
use these maps to predict responses to held-out questions
(Fig.~\ref{fig:predictions}) also has important psychological implications. One
such psychological implication is that concepts that are assigned to nearby
coordinates by the text embedding model also appear to be ``known to a similar
extent'' (as reflected by participants' responses to held-out questions). This
suggests that participants also \textit{conceptualize} similarly the content
reflected by nearby embedding coordinates. A second psychological implication
is that something about how participants' knowledge ``spreads'' across concepts
is being captured by the knowledge maps we are inferring from their quiz
responses (e.g., Figs.~\ref{fig:smoothness},~\ref{fig:knowledge-maps}). In
other words, our study shows that knowledge about a given concept implies
knowledge about related concepts.

In our study, we characterize the ``coordinates'' of participants' knowledge
using a relatively simple ``bag of words'' text embedding model~\citep[LDA;
]{BleiEtal03}. More sophisticated text embedding models, such as
transformer-based models~\citep{ViswEtal17, DevlEtal18, ChatGPT, TouvEtal23}
can learn complex grammatical and semantic relationships between words,
higher-order syntactic structures, sylistic features, and more. We considered
using transformer-based models in our study, but we were surprised to find that
the text embeddings derived from these models were surprisingly uninformative
with respect to differentiating or otherwise characterizing the conceptual
content of the lectures and questions we used. We suspect that this reflects a
broader challenge in constructing models that are high-resolution within a
given domain (e.g., the domain of physics lectures and questions) \textit{and}
sufficiently broad so as to enable them to cover a wide range of domains. For
example, we found that the embeddings derived even from much larger and more
modern models like BERT~\citep{DevlEtal18}, GPT~\citep{ViswEtal17},
LLaMa~\citep{TouvEtal23}, and others that are trained on enormous text corpora,
end up yielding poor resolution within the content space spanned by indiviudal
course videos (Supp. Fig.~\ldaVsBERT). Whereas the LDA embeddings of the
lectures and questions are ``near'' each other (i.e., the convex hull enclosing
the two lectures' trajectories is highly overlapping with the convex hull
enclosing the questions' embeddings), the BERT embeddings of the lectures and
questions are instead largely distinct (top row of Supp. Fig.~\ldaVsBERT). The
LDA embeddings of the questions for each lecture and the corresponding
lecture's trajectory are also similar. For example, as shown in
Fig.~\ref{fig:sliding-windows}C, the LDA embeddings for \textit{Four
Fundamental Forces} questions (blue dots) appear closer to the \textit{Four
Fundamental Forces} lecture trajectory (blue line), whereas the LDA embeddings
for \textit{Birth of Stars} questions (green dots) appear closer to the
\textit{Birth of Stars} lecture trajectory (green line). The BERT embeddings of
the lectures and questions do not show this property (Supp. Fig.~\ldaVsBERT).
We also examined per-question ``content matches'' between individual questions
and individual moments of each lecture
(Figs.~\ref{fig:question-correlations},~\ldaVsBERT). The timeseries plot of
individual questions' correlations are different from each other when computed
using LDA (e.g., the traces can be clearly visually separated), whereas the
correlations computed from BERT embeddings of different questions all look very
similar. This tells us that LDA is capturing some differences in content
between the questions, whereas BERT is not. The timeseries plots of individual
questions' correlations have clear ``peaks'' when computed using LDA, but not
when computed using BERT. This tells us that LDA is capturing a ``match''
between the content of each question and a relatively well-defined time window
of the corresponding lectures. The BERT embeddings appear to blur together the
content of the questions versus specific moments of each lecture. Finally, we
also compared the pairwise correlations between embeddings of questions within
versus across content areas (i.e., content covered by the individual lectures,
lecture-specific questions, and by the ``general physics knowledge''
questions). The LDA embeddings show a strong contrast between same-content
embeddings versus across-content embeddings. In other words, the embeddings of
questions about the \textit{Four Fundamental Forces} material are highly
correlated with the embeddings of the \textit{Four Fundamental Forces} lecture,
but not with the embeddings of \textit{Birth of Stars}, questions about
\textit{Birth of Stars}, or general physics knowledge questions. We see a
similar pattern with the LDA embeddings of the \textit{Birth of Stars}
questions (Fig.~\ref{fig:topics}, Supp. Fig.~\topicWeights). In contrast, the
BERT embeddings are all highly correlated with each other (Supp.
Fig.~\ldaVsBERT). Taken together, these comparisons illustrate how LDA (trained
on the specific content in question) provides both coverage of the requisite
material and specificity at the level of the content covered by individual
questions. BERT, on the other hand, essentially assigns both lectures and all
of the questions (which are all broadly about ``physics'') into a tiny region
of its embedding space, thereby blurring out meaningful distinctions between
different specific concepts covered by the lectures and questions. We note that
these are not criticisms of BERT (or other large language models trained on
large and diverse corpora). Rather, our point is that simple fine-tuned models
trained on a relatively small but specialized corpus can outperform much more
complicated models trained on much larger corpora, when we are specifically
interested in capturing subtle conceptual differences at the level of a single
course lecture or question. Of course if our goal had been to find a model that
generalized to many different content areas, we would expect our approach to
perform comparatively poorly relative to BERT or other much larger models. We
suggest that bridging the tradeoff between high resolution within each content
area versus the ability to generalize to many different content areas will be
an important challenge for future work in this domain.

Another application for large language models that does \textit{not} require
explicitly modeling the content of individual lectures or questions is to
leverage the models' ability to generate text. For example, generative text
models like ChatGPT~\citep{ChatGPT} and LLaMa~\citep{TouvEtal23} are already
being used to build a new generation of interactive tutoring
systems~\citep[e.g.,][]{MannEtal23b}.  Unlike the approach we have taken here,
these systems do not explicitly model what learners know, or how their knowledge 
changes over time with training.  One could imagine building a hybrid system
that combines the best of both worlds: a large language model that can
\textit{generate} text, combined with a smaller model that can \textit{infer}
what learners know and how their knowledge changes over time.  Such a hybrid
system could potentially be used to build a new generation of interactive
tutoring systems that are able to adapt to learners' needs in real time, and
that are able to provide more nuanced feedback about what learners know and
what they do not know.

At the opposite end of the spectrum from large language models, one could also
imagine \textit{simplifying} some aspects of our LDA-based approach by
computing simple word overlap metrics. For example, the Jaccard similarity
between text $A$ and $B$ is computed as the number of unique words in the
intersection of words from $A$ and $B$ divided by the number of unique words in
the union of words from $A$ and $B$. In a supplemental analysis (Supp.
Fig.~\jaccard), we compared the LDA-based question-lecture matches we reported
in Figure~\ref{fig:question-correlations} with the Jaccard similarities between
each question and each sliding window of text from the corresponding lecture.
As shown in Supplementary Figure~\jaccard, this simple word-matching approach
does not appear to capture the same level of specificity as the LDA-based
approach. For example, whereas the LDA-based approach often yields a clear peak
in the timeseries of correlations between each question and the corresponding
lecture, the Jaccard similarity-based approach does not. Furthermore, these
LDA-based matches appear to capture conceptual overlaps between the questions
and lectures (Supp. Tab.~\matchTab), whereas simple word matching does not. For
example, one of the example questions examined in Supplementary Figure~\jaccard
asks ``Which of the following occurs as a cloud of atoms gets more dense?''.
The LDA-based matches identify lecture timepoints where the releveant
\textit{topics} are discussed (e.g., when words like ``cloud,'' ``atom,''
``dense,'' etc., are mentioned \textit{together}). The Jaccard similarity-based
matches, on the other hand, are strong when \textit{any} of these words are
mentioned, even if they do not occur together.

We view our approach as occupying a sort of ``sweet spot,'' between much larger
language models and simple word matching-based approaches, that enables us to
capture the relevant conceptual content of course materials at an appropriate
semantic scale. Our approach enables us to accurately and consistently identify
each question's content in a way that also matches up with what is presented in
the lectures. In turn, this enables us to construct accurate predictions about
participants' knowledge of the conceptual content tested by held-out questions
(Fig.~\ref{fig:predictions}).

One limitation of our approach is that topic models contain no explicit
internal representations of more complex aspects of ``knowledge,'' like
knowledge graphs, dependencies or associations between concepts, causality, and
so on. These representations might (in principle) be added as extensions to our
approach to more accurately and precisely capture, characterize, and track
learners' knowledge. However, modeling these aspects of knowledge will likely
require substantial additional research effort.

@@

We hand-selected two course videos from the Khan Academy platform: \textit{Four
Fundamental Forces} (an introduction to gravity, electromagnetism, the weak
nuclear force, and the strong nuclear force; duration: 10 minutes and 29
seconds) and \textit{Birth of Stars} (an introduction to how stars are formed;
duration: 7 minutes and 57 seconds). All participants viewed the videos in the
same order (i.e., \textit{Four Fundamental Forces} followed by \textit{Birth of
Stars}). While we are not aware of any specific confounds of viewing order, nor
have are we aware of how or why viewing order might influence our main
findings, we acknowledge that we did not control for potential order effects in
our study.

We then hand-created 39 multiple-choice questions: 15 about the conceptual
content of \textit{Four Fundamental Forces} (i.e., Lecture~1), 15 about the
conceptual content of \textit{Birth of Stars} (i.e., Lecture~2), and 9
questions that tested for general conceptual knowledge about basic physics
(covering material that was not presented in either video). One of our group's
undergraduate research assistants worked alongside a rotating Masters student
to develop this set of questions (these researchers are acknowledged in our
paper for their contribution, although they did not meet the criteria for
authorship discussed with all team members at the start of the project, as
determined by J.R.M.) The senior author (J.R.M.) tasked the pair of researchers
with coming up with ``15 conceptual questions about each lecture, along with 9
additional questions about general physics knowledge.'' To help broaden the set
of lecture-specific questions, the researchers were further instructed to work
through each lecture in small segments, identify what each segment was
``about'' conceptually, and then write a question about that concept. The
general physics questions were drawn from the researchers' coursework along
with internet searches and brainstorming with the project team and other
members of J.R.M.'s lab. The final set of questions (and response options) was
reviewed and approved by J.R.M. before we collected or analyzed the text or
experimental data.

We note that estimating the specific ``amount'' of conceptual understanding
that each question ``requires'' to answer is somewhat subjective, and might
even come down to the ``strategy'' a given participant uses to answer the
question at that particular moment. The full set of questions and answer
choices may be found in Supplementary Table~\questions.

@@

Intuitively, $\mathrm{ncorr}(x, y)$ is the correlation between two topic
vectors (e.g., the topic vector from one timepoint in a lecture, $x$, and the
topic vector for one question, $y$), normalized by the minimum and maximum
correlations (across all timepoints $t$ and questions $Q$) to range between 0
and 1, inclusive. Equation~\ref{eqn:prop} then computes the weighted average
proportion of correctly answered questions about the content presented at
timepoint $t$, where the weights are given by the normalized correlations
between timepoint $t$'s topic vector and the topic vectors for each question.
The normalization step (i.e., using $\mathrm{ncorr}$ instead of the raw
correlations) ensures that every question contributes some non-negative amount
to the knowledge estimate.

>>>>

% supplemental tables
\newcommand{\questions}{1}
\newcommand{\topics}{2}

% supplemental figures
\newcommand{\topicWordWeights}{1}
\newcommand{\topicWeights}{2}
\newcommand{\individualKnowledgeMapsA}{3}
\newcommand{\individualKnowledgeMapsB}{4}
\newcommand{\individualKnowledgeMapsC}{5}
\newcommand{\individualLearningMapsA}{6}
\newcommand{\individualLearningMapsB}{7}

@@

Designing and building procedures and tools for mapping out knowledge touches
on deep questions about what it means to learn. For example, how do we acquire
conceptual knowledge? Memorizing course lectures or textbook chapters by rote
can lead to the superficial \textit{appearance} of understanding the underlying
content, but achieving true conceptual understanding seems to require something
deeper and richer. Does conceptual understanding entail connecting newly
acquired information to the scaffolding of one's existing knowledge or
experience~\citep{BlayEtal06,CaraMaho03, ConsEtal16, DeacEtal04, SimoEtal04}?
Or weaving a lecture's atomic elements (e.g., its component words) into a
structured network that describes how those individual elements are
related~\citep{LeeChen22}? Conceptual understanding could also involve building
a mental model that transcends the meanings of those individual atomic elements
by reflecting the deeper meaning underlying the gestalt whole~\citep{Kint70,
Macl05, ScotEtal07}.

The difference between ``understanding'' and ``memorizing,'' as framed by
researchers in education, cognitive psychology, and cognitive
neuroscience (e.g.,~\citealp{Kato40, Gall00, ScotEtal07, HallGree08, Macl05}),
has profound analogs in the fields of natural language processing and natural
language understanding. For example, considering the raw contents of a document
(e.g., its constituent symbols, letters, and words) might provide some
clues as to what the document is about, just as memorizing a passage
might provide some ability to answer simple questions about it. However, text
embedding models (e.g.,~\citealp{LandDuma97, DeerEtal90, BleiEtal03,
BleiLaff06, MikoEtal13a, CerEtal18, BrowEtal20}) also attempt to capture the
deeper meaning \textit{underlying} those atomic elements. These models consider
not only the co-occurrences of those elements within and across documents, but
also patterns in how those elements appear across different scales (e.g.,
sentences, paragraphs, chapters, etc.), the temporal and grammatical properties
of the elements, and other high-level characteristics of how they are
used~\citep{Mann20, Mann21a}. According to these models, the deep conceptual
meaning of a document may be captured by a feature vector in a high-dimensional
representation space, wherein nearby vectors reflect conceptually related
documents. A model that succeeds at capturing an analogue of ``understanding'' is
able to assign nearby feature vectors to two conceptually related documents,
\textit{even when the specific words contained in those documents have very little
overlap}.

Given these insights, what form might a representation of the sum total of a
person's knowledge take? First, we might require a means of systematically
describing or representing the nearly infinite set of possible things a person
could know. Second, we might want to account for potential associations between
different concepts. For example, the concepts of ``fish'' and ``water'' might
be associated in the sense that fish live in water. Third, knowledge may have a
critical dependency structure, such that knowing about a particular concept
might require first knowing about a set of other concepts. For example,
understanding the concept of a fish swimming in water first requires
understanding what fish and water \textit{are}. Fourth, as we learn, our
``current state of knowledge'' should change accordingly. Learning new concepts
should both update our characterizations of ``what is known'' and also
unlock any now-satisfied dependencies of those newly learned concepts so that
they are ``tagged'' as available for future learning.

@@

@@

    \caption{\textbf{Estimating moment-by-moment knowledge acquisition.}
    \textbf{A. Moment-by-moment knowledge about the \textit{Four Fundamental
    Forces}}. Each trace displays the weighted proportion of correctly answered
    questions about the content reflected in each moment of the lecture (see
    \nameref{subsec:traces}), using responses from a single
    quiz (color). The traces are averaged across participants. \textbf{B.
    Average estimated knowledge about the \textbf{Four Fundamental Forces}.}
    Each bar displays the across-timepoint average knowledge, estimated using
    the responses to one quiz's questions. \textbf{C. Moment-by-moment
    knowledge about the \textit{Birth of Stars}}. The panel is in the same
    format as Panel A, but here the knowledge estimates are for the
    moment-by-moment content of the \textit{Birth of Stars} lecture. \textbf{D.
    Average estimated knowledge about the \textit{Birth of Stars}.} The panel
    is in the same format as Panel B, but here the knowledge estimates are for
    the content of the \textit{Birth of Stars} lecture. All panels: error
    ribbons and error bars denote 95\% confidence intervals, estimated across
    participants.}

@@

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should carry some
predictive information about whether the participant is likely to answer that
question correctly or incorrectly. We developed a statistical approach to test this claim.
For each question, in turn, we used Equation~\ref{eqn:prop} to estimate each
participant's knowledge at the given question's embedding space coordinate,
using all \textit{other} questions that participant answered on the same quiz.
For each quiz, we grouped these estimates into two distributions: one for the
estimated knowledge at the coordinates of \textit{correctly} answered
questions, and another for the estimated knowledge at the coordinates of
\textit{incorrectly} answered questions (Fig.~\ref{fig:predictions}). We then
used independent samples $t$-tests to compare the means of these distributions
of estimated knowledge.

\begin{figure}[tp]
    \centering
    \includegraphics[width=\textwidth]{figs/held-out-question-analysis}

    \caption{\textbf{Estimating knowledge at the embedding coordinates of
    held-out questions.} Separately for each quiz (panel), we plot the
    distributions of predicted knowledge at the embedding coordinates of each
    held-out correctly (blue) or incorrectly (red) answered question. The
    $t$-tests reported in each panel are between the distributions of estimated
    knowledge at the coordinates of correctly versus incorrectly answered
    held-out questions.}

@@

For the initial quizzes participants took (prior to watching either lecture),
participants' estimated knowledge tended to be low overall, and relatively
unstructured (Fig.~\ref{fig:predictions}, left panel). When we held out
individual questions and estimated their knowledge at the held-out questions'
embedding coordinates, we found no reliable differences in the estimates when
the held-out question had been correctly versus incorrectly answered ($t(633) =
0.577,~p = 0.564$). This reflects a floor effect: when knowledge is low
everywhere, there is little signal to differentiate between what is known
versus unknown. After watching the first lecture, estimated knowledge for
held-out correctly answered questions (from the second quiz;
Fig.~\ref{fig:predictions}, middle panel) exhibited a positive shift relative
to held-out incorrectly answered questions ($t(633) = 3.961,~p < 0.001$). This
second quiz provides the maximally sensitive test for our knowledge
predictions, since (if knowledge is estimated accurately) participants' Quiz~2
responses should demonstrate specific knowledge about Lecture~1 content, but
knowledge about Lecture~2 and general physics concepts should be roughly
unchanged from before they watched Lecture~1. After watching the second lecture,
estimated knowledge (from the third quiz; Fig.~\ref{fig:predictions}, right
panel) for \textit{all} questions exhibited a positive shift. However, the
estimated knowledge for held-out correctly answered questions remained
greater than that for held-out incorrectly answered questions ($t(628) = 2.045,~p
= 0.041$). This third contrast reflects a ceiling effect: when knowledge is
relatively high everywhere, the signal differentiating what is known versus
unknown is relatively weak. Taken together, this set of analyses demonstrates
that our knowledge prediction framework is most informative when participants
exhibit variability in their knowledge of the content captured by the text
embedding model.

Knowledge estimates need not be limited to the content of the lectures. As
illustrated in Figure~\ref{fig:knowledge-maps}, our general approach to
estimating knowledge from a small number of quiz questions may be extended to
\textit{any} content, given its text embedding coordinate. To visualize how
knowledge ``spreads'' through text embedding space to content beyond the
lectures participants watched, we first fit a new topic model to the lectures'
sliding windows with (up to)~$k = 100$~topics.
Conceptually, increasing the number of topics used by the model functions to
increase the ``resolution'' of the embedding space, providing a greater ability to
estimate knowledge for content that is highly similar to (but not precisely the same as)
that contained in the two lectures.
Aside from increasing the number of topics from 15 to 100, all other
procedures and model parameters were carried over from the preceding analyses.
As in our other analyses, we resampled each lecture's topic trajectory to 1~Hz
and projected each question into a shared text embedding space.

@@

@@

We hand-selected two course videos from the Khan Academy platform: \textit{Four
Fundamental Forces} (an introduction to gravity, electromagnetism, the weak
nuclear force, and the strong nuclear force; duration: 10 minutes and 29
seconds) and \textit{Birth of Stars} (an introduction to how stars are formed;
duration: 7 minutes and 57 seconds). We then hand-created 39 multiple-choice
questions: 15 about the conceptual content of \textit{Four Fundamental Forces} (i.e., Lecture~1),
15 about the conceptual content of \textit{Birth of Stars} (i.e., Lecture~2), and 9
questions that tested for general conceptual knowledge about basic physics
(covering material that was not presented in either video). The full set of
questions and answer choices may be found in Supplementary Table~\questions.

@@

Intuitively, $\mathrm{ncorr}(x, y)$ is the correlation between two topic
vectors (e.g., the topic vector from one timepoint in a lecture, $x$, and the
topic vector for one question, $y$), normalized by the minimum and maximum
correlations (across all timepoints $t$ and questions $Q$) to range between 0 and 1,
inclusive. Equation~\ref{eqn:prop} then computes the weighted average
proportion of correctly answered questions about the content presented at
timepoint $t$, where the weights are given by the normalized correlations
between timepoint $t$'s topic vector and the topic vectors for each question.
The normalization step (i.e., using $\mathrm{ncorr}$ instead of the raw
correlations) insures that every question
contributes some non-negative amount to the knowledge estimate.
